COATi: statistical pairwise alignment of protein coding sequences

### 2024-01-05 alignpair_letter.tex
Figure and supplement work

There are well-established algorithms for combining FSTs in different ways allowing the design of complex models by combining simpler FSTs, including concatenation, composition, intersection, union, and reversal \citep{bradley2007transducers,silvestre2021machine}. Specifically, composition is an algorithm to combine two FSTs by sending the output of one FST into the input of another, creating a new, more complex transducer. Figure~\ref{fig:base-calling} illustrates how FSTs modeling sequencing errors (Fig.~\ref{fig:base-calling}a) and ambiguity (Fig.~\ref{fig:base-calling}b) can be combined via composition to produce an FST that does both (Fig.~\ref{fig:base-calling}c). Conceptually, composition creates an FST that generates a descendant sequence from a known ancestor via an unknown intermediate, $J$, and can calculate the conditional probability $P(D|A) = \sum_{J} P(D|J)P(J|A)$.

@@

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials Figure 2.

>>>>

There are well-established algorithms for combining FSTs in different ways allowing the design of complex models by combining simpler FSTs, including concatenation, composition, intersection, union, and reversal \citep{bradley2007transducers,silvestre2021machine}. Specifically, composition is an algorithm to combine two FSTs by sending the output of one FST into the input of another, creating a new, more complex transducer \citep{mohri2005weighted}. Figure~\ref{fig:base-calling} illustrates how FSTs modeling sequencing errors (Fig.~\ref{fig:base-calling}a) and ambiguity (Fig.~\ref{fig:base-calling}b) can be combined via composition to produce an FST that does both (Fig.~\ref{fig:base-calling}c). Conceptually, composition creates an FST that generates a descendant sequence from a known ancestor via an unknown intermediate, $J$, and can calculate the conditional probability $P(D|A) = \sum_{J} P(D|J)P(J|A)$.

@@

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials Figure 1.

### 2024-01-05 supplementary_materials/supplementary_materials.Rmd
Figure and supplement work

---
title: "Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\""
author: Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright
output:
    pdf_document:
        fig_caption: yes
        keep_tex: true
bibliography: alignpair_letter.bib
---

@@

    ###########################################################################
    # perfect, best, and imperfect alignments
    source("supplementary_materials/scripts/number_alignments.R")
    alns = num_alns(results_file)

@@

    ############################################################################
    # selection (ka ks)
    source("supplementary_materials/scripts/kaks.R")

@@

@@

results = "supplementary_data/tri-mg/results/results_summary.csv"

@@

results = "supplementary_data/tri-ecm/results/results_summary.csv"

@@

results = "supplementary_data/mar-mg/results/results_summary.csv"

@@

results = "supplementary_data/mar-ecm/results/results_summary.csv"

@@

results = "supplementary_data/reverse/results/results_summary.csv"

@@

source("supplementary_materials/scripts/results_gor_ref.R")

@@

```{r base-calling-error, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap="\\label{fig:figs} Base calling error FST. Arcs from M to S generate matches; however, here they can introduce single-nucleotide errors, which can generate stop codon artifacts."}
knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
```

```{r aln-example, echo = FALSE, fig.show='hold', fig.align='center', fig.cap="\\label{fig:aln-example} FST of all possible alignments between `CTC' and `CTG' (top) and the best alignment, determined by the shortest path (bottom). Figures generated using the OpenFST library."}
knitr::include_graphics(path = c("figures/aln-example-graph.pdf", "figures/aln-example.pdf"))
```

```{r pvals, echo = FALSE}
results = "supplementary_data/tri-mg/results/results_summary.csv"
pval_tri_mg = max(wilcoxon_test(results, "tri-mg"))
results = "supplementary_data/tri-ecm/results/results_summary.csv"
pval_tri_ecm = max(wilcoxon_test(results, "tri-ecm"))
results = "supplementary_data/mar-mg/results/results_summary.csv"
pval_mar_mg = max(wilcoxon_test(results, "mar-mg"))
results = "supplementary_data/mar-ecm/results/results_summary.csv"
pval_mar_ecm = max(wilcoxon_test(results, "mar-ecm"))
results = "supplementary_data/reverse/results/results_summary.csv"
pval_reverse = max(wilcoxon_test(results, "tri-mg"))
```

```{r dseq1, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/tri-mg/plot_distance.csv", "tri-mg", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_tri-mg.pdf")
```

```{r dseq2, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/tri-ecm/plot_distance.csv", "tri-ecm", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_tri-ecm.pdf")
```

```{r dseq3, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/mar-mg/plot_distance.csv", "mar-mg", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_mar-mg.pdf")
```

```{r dseq4, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/mar-ecm/plot_distance.csv", "mar-ecm", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_mar-ecm.pdf")
```

```{r dseq5, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-reverse} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE with gorilla as the reference. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_reverse, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/reverse/plot_distance.csv", "reverse", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_reverse.pdf")
```

@@

```{r k2p-dist, echo = FALSE, fig.cap="\\label{fig:k2p} K2P distances. Evolutionary distances between reference and retrieved alignments for each method calculated using the Kimura 2-parameter model. Notably, COATi performs best while the other tools tend to overestimate the evolutionary distance."}
knitr::include_graphics(path = "figures/k2p_distances.pdf")
```

>>>>

---
title: "COATi: statistical pairwise alignment of protein coding sequences"
subtitle: Supplementary Materials
author: Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright
mainfont: TeX Gyre Pagella
bibliography: ../alignpair_letter.bib
output:
    "bookdown::pdf_document2":
        numbered_sections: yes
        fig_caption: yes
        keep_tex: true
        latex_engine: lualatex
header-includes:
  - \let\oldthefigure\thefigure
  - \renewcommand{\thefigure}{S\oldthefigure}
---

@@

    ###########################################################################
    # perfect, best, and imperfect alignments
    source("scripts/number_alignments.R")
    alns = num_alns(results_file)

@@

    ############################################################################
    # selection (ka ks)
    source("scripts/kaks.R")

@@

# Aligner Commands

We evaluated five different aligners. Below are the commands that we used to run them. We have abbreviated the commands for clarity, stripping out unimportant arguments. Complete workflows can be found in our coati-testing repository on Github.

- COATi: `coati alignpair -m tri-mg ...`
- ClustalΩ v1.2.4: `clustalo ...`
- PRANK v.150803: `prank -codon ...`
- MACSE v2.06: `java -jar macse.jar -fs_lr 10 -stop_lr 10 -prog alignSequences ...`
- MAFFT v7.505: `mafft --preservecase --globalpair --maxiterate 1000 ...`

COATi can use different alignment models for pairwise alignment. Below are the commands that we used to run different models.

- tri-mg: `coati alignpair -m tri-mg ...`
- tri-ecm: `coati alignpair -m tri-ecm ...`
- mar-mg: `coati alignpair -m mar-mg ...`
- mar-ecm: `coati alignpair -m mar-ecm ...`

# FST Alignment Example

When using the triplet models (tri-mg and tri-ecm), COATi uses the OpenFST library to generate best alignments by composing the input and output sequences with the COATi FST model. While the COATi FST model is too large to display, we can show the result of a composition.

Fig. \@ref(fig:aln-example-a) shows a graph depicting the FST that results from composing the COATi FST with the input sequence "CTC" and the output sequence "CTG". Every path through this FST represents one possible way to align "CTC" and "CTG", and the sum of all weights along a path is the total weight of the respective alignment. Here, the weights of each arc are in negative-log space. Note that this graph has been optimized, and weight has been pushed towards the initial state. The weight of any specific arc may not be directly mapable to a weight described in the model.

Fig. \@ref(fig:aln-example-b) is the best alignment between "CTC" and "CTG", as determined by the shortest path algorithm.  Figs. \@ref(fig:aln-example-a) and \@ref(fig:aln-example-b) were produced by the OpenFST library. A bold circle represents a starting node, and a double circle represents a termination node.

```{r aln-example-a, echo = FALSE, fig.align='center', fig.cap="The FST of all possible alignments between `CTC' and `CTG'."}
knitr::include_graphics("figures/aln-example-graph.pdf")
```

```{r aln-example-b, echo = FALSE, out.width="75%", fig.align='center', fig.cap="The best alignment of `CTC' and `CTG'."}
knitr::include_graphics("figures/aln-example-path.pdf")
```

@@

results = "data/tri-mg/results/results_summary.csv"

@@

results = "data/tri-ecm/results/results_summary.csv"

@@

results = "data/mar-mg/results/results_summary.csv"

@@

results = "data/mar-ecm/results/results_summary.csv"

@@

results = "data/reverse/results/results_summary.csv"

@@

source("scripts/results_gor_ref.R")

@@

```{r base-calling-error, echo = FALSE, fig.width = 6, fig.height = 6, fig.cap="\\label{fig:figs} Base calling error FST. Arcs from M to S generate matches; however, here they can introduce single-nucleotide errors, which can generate stop codon artifacts."}
#knitr::include_graphics(path = "figures/fig-base-calling-error.pdf")
```

```{r pvals, echo = FALSE}
results = "data/tri-mg/results/results_summary.csv"
pval_tri_mg = max(wilcoxon_test(results, "tri-mg"))
results = "data/tri-ecm/results/results_summary.csv"
pval_tri_ecm = max(wilcoxon_test(results, "tri-ecm"))
results = "data/mar-mg/results/results_summary.csv"
pval_mar_mg = max(wilcoxon_test(results, "mar-mg"))
results = "data/mar-ecm/results/results_summary.csv"
pval_mar_ecm = max(wilcoxon_test(results, "mar-ecm"))
results = "data/reverse/results/results_summary.csv"
pval_reverse = max(wilcoxon_test(results, "tri-mg"))
```

```{r dseq1, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
library(knitr)
source("scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("data/tri-mg/plot_distance.csv", "tri-mg", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_tri-mg.pdf")
```

```{r dseq2, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
library(knitr)
source("scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("data/tri-ecm/plot_distance.csv", "tri-ecm", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_tri-ecm.pdf")
```

```{r dseq3, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
library(knitr)
source("scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("data/mar-mg/plot_distance.csv", "mar-mg", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_mar-mg.pdf")
```

```{r dseq4, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
library(knitr)
source("scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("data/mar-ecm/plot_distance.csv", "mar-ecm", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_mar-ecm.pdf")
```

```{r dseq5, echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-reverse} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE with gorilla as the reference. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_reverse, digits = 3), "$.")}
library(knitr)
source("scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("data/reverse/plot_distance.csv", "reverse", save = TRUE))
knitr::include_graphics(path = "figures/dseq_plots_reverse.pdf")
```

@@

```{r k2p-dist, echo = FALSE, fig.align = 'center', out.width="75%", fig.cap="\\label{fig:k2p} Alignments generated by COATi produce accurate evolutionary distances. Benchmark alignments were realigned by COATi and four other methods as described in the main text. Kimura two-parameter (K2P) distances were estimated from the benchmark alignments and the estimated alignments. Each panel is a scatter plot comparing the benchmark disances (Reference) with distances estimated via one of the aligners. COATi clearly performs better than other tools, which tend to drastically overestimate some distances."}
knitr::include_graphics(path = "figures/k2p_distances.pdf")
```
